---
title: '**Linear models, functions, shortcuts**'
author: "Kevin Aagaard"
date: "November 16, 2015"
output: pdf_document
---

```{r, echo=F}
#require(data.table)
#require(plyr)
#require(...)

x = c("data.table", "plyr", "readr", "readxl", "microbenchmark")
lapply(x,require,character.only=T)
```

# **Linear Models**

**The command**:

`?lm()`
 
>  lm is used to fit linear models. It can be used to carry out regression, single stratum analysis of variance and 
analysis of covariance (although aov may provide a more 
convenient interface for these).

```{r} 
#lm(formula, data, subset, weights, na.action, method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE, singular.ok = TRUE, contrasts = NULL, offset, ...)
```


**formula**: an object of class "formula" (or one that can be 
         coerced to that class): a symbolic description of 
         the model to be fitted. The details of model 
         specification are given under 'Details'.  
**data**:    an optional data frame, list or environment (or 
         object coercible by as.data.frame to a data frame) 
         containing the variables in the model. If not found 
         in data, the variables are taken from 
         environment(formula), typically the environment from 
         which lm is called.  
**subset**:  an optional vector specifying a subset of 
         observations to be used in the fitting process.  
**weights**: an optional vector of weights to be used in the 
         fitting process. Should be NULL or a numeric vector. 
         If non-NULL, weighted least squares is used with 
         weights weights (that is, minimizing sum(w*e^2)); 
         otherwise ordinary least squares is used. See also 
         'Details'.  
**na.action**:  a function which indicates what should happen 
            when the data contain NAs. The default is set by 
            the na.action setting of options, and is na.fail 
            if that is unset. The 'factory-fresh' default is 
            na.omit. Another possible value is NULL, no 
            action. Value na.exclude can be useful.  
**method**:  the method to be used; for fitting, currently only 
         method = "qr" is supported; method = "model.frame" 
         returns the model frame (the same as with model = 
         TRUE, see below).  
**model, x, y, qr**:  logicals. If TRUE the corresponding 
                  components of the fit (the model frame, 
                  the model matrix, the response, the QR 
                  decomposition) are returned.  
**singular.ok**:   logical. If FALSE (the default in S but not 
               in R) a singular fit is an error.  
**contrasts**:   an optional list. See the contrasts.arg of 
             model.matrix.default.  
**offset**:   this can be used to specify an a priori known 
          component to be included in the linear predictor 
          during fitting. This should be NULL or a numeric 
          vector of length equal to the number of cases. One 
          or more offset terms can be included in the 
          formula instead or as well, and if more than one 
          are specified their sum is used. See model.offset.  
**...**:      additional arguments to be passed to the low level 
          regression fitting functions (see below).  

## **Example**
```{r}
tree_data = read.table("Trees.txt", header=T, sep="\t")
```
*Formula (is DBH related to Species?)*: DBH ~ index(BOTANICAL)  
*Data*: Trees  

```{r}
tree_data$species_id = id(tree_data["BOTANICAL"], drop=T)

example_lm = lm(DBH ~ species_id, data=tree_data)

#Output
example_lm

summary(example_lm)
```

### *Regression plot*  
```{r}
plot(x=tree_data$species_id,y=tree_data$DBH,type="p")
abline(example_lm)
abline(a=47.8526, b = 0.1987, col="red", lwd=5, lty=2)
```

### *QQ plots*

```{r} 
plot(example_lm)
```

1) Possible non-linear trend  
2) Possibly non-normal (or, "non-Gaussian") residuals  
3) Potential non-constant error variance)  
4) Trend may be driven by a single point  

***

## **Multiple responses**  
```{r}
example_lm2 = lm(DBH ~ species_id+SITE, data=tree_data)

#Output
example_lm2

summary(example_lm2)
```

### *QQ plots*

```{r}
plot(example_lm2)
```

1) Possible non-linear trend  
2) Possibly non-normal (or, "non-Gaussian") residuals  
3) Potential non-constant error variance)  
4) Low leverage, high resisual  

***

# **Analysis of Variance**

**The command**:

`?aov()`

> Fit an analysis of variance model by a call to lm for each stratum.

```{r}
#aov(formula, data = NULL, projections = FALSE, qr = TRUE, contrasts = NULL, ...)
```

**formula**:   A formula specifying the model.  
**data**:      A data frame in which the variables specified in 
           the formula will be found. If missing, the 
           variables are searched for in the standard way.  
**projections**:  Logical flag: should the projections be returned?  
**qr**:           Logical flag: should the QR decomposition be 
              returned?  
**contrasts**:    A list of contrasts to be used for some of the 
              factors in the formula. These are not used for 
              any Error term, and supplying contrasts for 
              factors only in the Error term will give a 
              warning.  
**...**:      Arguments to be passed to lm, such as subset or 
          na.action. See 'Details' about weights.  

*Formula (is DBH related to Species?)*: DBH ~ index(BOTANICAL)  
*Data*: Trees  

```{r,echo=F}
example_aov = aov(example_lm)

#Output
example_aov

summary(example_aov)
```


## **Multiple responses**  
```{r,echo=F}
example_aov2 = aov(example_lm2)

#Output
example_aov2

summary(example_aov2)
```

***

# **Functions**

##  **Why use functions?**  

### **For speed!**  

*To time stuff*: proc.time()  

```{r, echo=T, eval=F}
test = rnorm(1000, 10, 1)

z_test = vector()

timer = proc.time()
for(i in 1:length(test)) {
  z_test[i] = (test[i] - mean(test)) / sd(test)
}
proc.time() - timer
#> proc.time() - timer
#user  system elapsed 
#81.78    4.81   88.08 

zscore.func = function(x) (x - mean(x))/sd(x)
timer = proc.time()
z_test_fun = zscore.func(test)
proc.time() - timer
#> proc.time() - timer
#user  system elapsed 
#0.0     0.0     0.9 
```


### **For flexibility! (Not everything has been made for R...)**
```{r}
bird_data = read.table("Waterfowl.txt", header=T, sep="\t")

bird_data$Observed_Date = as.Date(bird_data$Observed_Date,
                                  "%m/%d/%Y")

#Bird-use-days function
BUDS = function(x, day) {
  birdusedays = array(NA, dim=c(dim(x)[1],dim(x)[2]))
  for (j in 1:dim(x)[1]) {
    for (k in 2:dim(x)[2]) {
      birdusedays[j,k] = (day[k]-day[k-1])*((x[j,k]+x[j,k-1])/2)
    } # k
  } # j
  results = rowSums(birdusedays, na.rm=TRUE)
  return(round(results))
}

bird_data_bud = 
  dcast.data.table(data.table(bird_data),
                   Waterbody_ID~Observed_Date, fun.aggregate=sum,
                   fill=0, value.var="Observed")

bird_data_days = 
  as.numeric(round(difftime(colnames(bird_data_bud)[-1],
                            min(colnames(bird_data_bud)[-1]), 
                            units="days"))) + 1

waterfowl_buds = BUDS(data.frame(bird_data_bud)[,-1], 
                      bird_data_days)

waterfowl_buds = data.table(waterfowl_buds)
waterfowl_buds
```

***

# **Shortcuts**

## **Loading multiple libraries at once**  

```{r}
#require(data.table)
#require(plyr)
#require(...)

x = c("data.table", "plyr", "readr", "readxl", "microbenchmark")
lapply(x,require,character.only=T)
```


## **Reading in large data files**
```{r, echo=T, eval=F}
data_test1 = read.csv("big_Trees.csv")
data_test2 = read_delim("big_Trees.csv", delim=",")
data_test3 = read_excel("big_Trees.xlsx")
```
```{r}
timer = proc.time()
data_test1 = read.csv("big_Trees.csv")
proc.time() - timer

timer = proc.time()
data_test2 = read_delim("big_Trees.csv", delim=",")
proc.time() - timer

timer = proc.time()
data_test3 = read_excel("big_Trees.xlsx")
proc.time() - timer
```


## **For loops, Vectorization, Functions**
```{r}
test = rnorm(1000000, 10, 1)

add_test = vector()

#timer = proc.time()
#for(i in 1:length(test)) {
#  add_test[i] = test[i] + 1
#}
#proc.time() - timer
#> proc.time() - timer
#user  system elapsed 
#1837.85  494.32 2394.03 <- 39 minutes!

add.func = function(x) x+1
timer = proc.time()
add_test_fun = add.func(test)
proc.time() - timer

timer = proc.time()
add_test_vector = test + 1
proc.time() - timer
```


## **Dynamically saving tables and graphs**
```{r}
#for(
#  i in 1:#NUMBER OF FILES TO SAVE
#    ) 
#  {
#  filename = paste(names(
#    #R OBJECTS TO SAVE
#    )[i], 
#    #" SOME EXTRA DESCRIPTOR OF EACH FILE", 
#    ".jpeg", sep="")
#  save_files_to = file.path(paste(getwd()), filename)
#  
#  jpeg(file = 
#        #R OBJECT
#        , width = 960, height = 480)
#  
#  plot(
#    #R OBJECT
#    [[i]], main = names(
#      #R OBJECT
#      )[i])
#  
#  dev.off()
#}
```


## **Easier subsetting with data.tables**
```{r}
subset_example = read_csv("Trees.csv")
subset_example_sites = subset_example[subset_example$SITE==1,]

#AND
subset_example_large = 
  subset_example[subset_example$SITE==1 &
                   subset_example$DBH>=50,]
subset_example_large

#OR
subset_example_species = 
  subset_example[subset_example$BOTANICAL=="Quercus alba" | 
                   subset_example$BOTANICAL=="Acer saccharinum",]
subset_example_species

#%in%

"stump" %in% subset_example$"COMMON NAME"
```


## **Doing stuff to every element of two arrays**
```{r}
array_test = c(1:10)
array_test1 = seq(0.1, 1, 0.1)

array_test - array_test1

cbind(array_test, array_test1, array_test - array_test1)

array_test2=vector("list")
for(i in 1:length(array_test)){
  array_test2[[i]] = array_test[i] - array_test1
}
array_test2

data_frame_test2 = data.frame(matrix(unlist(array_test2), 
                                     nrow = 10, byrow=T))
data_frame_test2

#What if array_test and array_test1 aren't the same length?

timer = proc.time()
array_test3 = outer(array_test, array_test1, `-`)
proc.time() - timer

timer = proc.time()
array_test4 = lapply(array_test, "-", array_test1)
proc.time() - timer

timer = proc.time()
x = rep(array_test, each=length(array_test))
array_test5 = split(x-array_test1, x)
proc.time() - timer

timer = proc.time()
array_test6 = CJ(array_test, array_test1)[, V:=V1-V2]
proc.time() - timer
```